# The Basic Genetic Model {#sec-genetic-model}

<div class="learning-objectives">
### Learning Objectives {.unnumbered}

By the end of this chapter, you will be able to:

1. Partition a phenotype into genetic and environmental components
2. Distinguish between additive, dominance, and epistatic genetic effects
3. Explain why additive effects are the primary focus of animal breeding
4. Understand the concept of variance partitioning
5. Calculate breeding values from a simple genetic model

</div>

```{r}
#| echo: false
#| message: false
#| warning: false

# Load required packages
library(tidyverse)
library(knitr)

# Set default theme for plots
theme_set(theme_minimal(base_size = 12))
```

## Introduction

[Content to be developed: Introduce the need for a genetic model to partition phenotypes into heritable and non-heritable components.]

## The Basic Model: y = μ + A + E

[Content to be developed: Present the fundamental genetic model.]

### Components of the Model

[Content to be developed: Define each component:]

- **y**: Phenotype (observed value)
- **μ**: Population mean
- **A**: Additive genetic effect (breeding value)
- **E**: Environmental deviation (everything else)

### Why This Model Matters

[Content to be developed: This simple model is the foundation of animal breeding. It allows us to separate genetic merit from environmental effects.]

## Types of Genetic Effects

[Content to be developed: Not all genetic effects are created equal.]

### Additive Genetic Effects

[Content to be developed:]

- Definition: Sum of average effects of alleles at all loci
- Breed true (passed from parent to offspring)
- Determine breeding value
- Primary target of selection

### Dominance Effects

[Content to be developed:]

- Definition: Within-locus interaction between alleles (heterozygote deviates from midparent)
- Do NOT breed true
- Not transmitted reliably to offspring
- Contribute to heterosis in crossbreeding

### Epistatic Effects

[Content to be developed:]

- Definition: Between-locus interactions
- Complex and difficult to predict
- May contribute to heterosis
- Not reliably transmitted

## Why Focus on Additive Effects?

[Content to be developed: Explain the primacy of additive effects in animal breeding:]

1. Additive effects are **predictably transmitted** from parent to offspring
2. **Response to selection** depends on additive genetic variance
3. **Breeding values** (what we estimate) are additive effects
4. Dominance and epistasis don't breed true

## Variance Partitioning

[Content to be developed: Total phenotypic variance can be partitioned into genetic and environmental components.]

### The Variance Equation

[Content to be developed:]

$$
\sigma^2_P = \sigma^2_G + \sigma^2_E
$$

Where:

- σ²_P = phenotypic variance
- σ²_G = genetic variance (total)
- σ²_E = environmental variance

### Further Partitioning Genetic Variance

[Content to be developed:]

$$
\sigma^2_G = \sigma^2_A + \sigma^2_D + \sigma^2_I
$$

Where:

- σ²_A = additive genetic variance
- σ²_D = dominance variance
- σ²_I = epistatic (interaction) variance

### Why Variance Matters

[Content to be developed: Variance components determine heritability (Chapter 5) and response to selection (Chapter 6).]

## Examples Across Species

[Content to be developed: Show the basic genetic model applied to real traits.]

### Example 1: Weaning Weight in Swine

[Content to be developed: Numerical example showing phenotype, genetic effect, and environmental deviation.]

### Example 2: Milk Yield in Dairy Cattle

[Content to be developed: Numerical example.]

### Example 3: Body Weight in Broilers

[Content to be developed: Numerical example.]

## Parent-Offspring Relationships

[Content to be developed: How additive effects create resemblance between relatives.]

### Offspring Breeding Value

[Content to be developed:]

$$
BV_{offspring} = \frac{1}{2} BV_{sire} + \frac{1}{2} BV_{dam} + Mendelian \ sampling
$$

### Mendelian Sampling

[Content to be developed: Offspring deviate from midparent average due to random sampling of parental alleles.]

## R Demonstration: Simulating the Basic Model

[Content to be developed: R code to simulate phenotypes from genetic and environmental components.]

```{r}
#| echo: true
#| eval: false

# Simulate phenotypes from genetic and environmental effects
set.seed(123)
n_animals <- 1000

# Population parameters
mu <- 100          # Population mean (e.g., kg for weaning weight)
sigma_A <- 5       # Additive genetic SD
sigma_E <- 10      # Environmental SD

# Simulate breeding values (additive genetic effects)
A <- rnorm(n_animals, mean = 0, sd = sigma_A)

# Simulate environmental deviations
E <- rnorm(n_animals, mean = 0, sd = sigma_E)

# Calculate phenotypes
y <- mu + A + E

# Summarize
data.frame(
  Component = c("Mean", "SD(A)", "SD(E)", "SD(y)"),
  Value = c(mean(y), sd(A), sd(E), sd(y))
)
```

## Summary

[Content to be developed: Recap main points.]

### Key Points

- Phenotype = Genetic + Environment (y = μ + A + E)
- Genetic effects include additive, dominance, and epistatic components
- Additive effects breed true and are the focus of selection
- Variance partitioning separates genetic from environmental variation
- Only additive genetic variance contributes to response to selection

## Practice Problems

[Problems to be developed]

1. A pig has a phenotype of 25 kg for weaning weight. The population mean is 23 kg, and the pig's breeding value is +3 kg. What is the environmental deviation for this pig?

2. Explain why dominance effects do not contribute to long-term response to selection.

3. If σ²_P = 100, σ²_A = 30, and σ²_E = 60, what is σ²_D + σ²_I?

## Further Reading

[References to be added]

- Falconer and Mackay: Introduction to Quantitative Genetics (Chapter 7)
- Lynch and Walsh: Genetics and Analysis of Quantitative Traits
