# Quantitative vs. Simply Inherited Traits {#sec-quantitative-traits}

<div class="learning-objectives">
### Learning Objectives {.unnumbered}

By the end of this chapter, you will be able to:

1. Distinguish between quantitative and simply inherited traits
2. Provide examples of each type in livestock species
3. Explain why most economically important traits are quantitative
4. Describe how breeding approaches differ for the two types of traits
5. Understand the concept of threshold traits

</div>

```{r}
#| echo: false
#| message: false
#| warning: false

# Load required packages
library(tidyverse)
library(knitr)

# Set default theme for plots
theme_set(theme_minimal(base_size = 12))
```

## Introduction

[Content to be developed: Not all traits inherit in the same way. Understanding the genetic architecture helps us choose appropriate breeding strategies.]

## Simply Inherited Traits

[Content to be developed: Traits controlled by one or a few genes with major effects.]

### Characteristics

[Content to be developed:]

- Mendelian inheritance (follows simple ratios)
- Qualitative expression (discrete categories)
- Major gene effects
- Can be eliminated or fixed through selection
- Molecular tests available for many

### Examples in Livestock

[Content to be developed:]

#### Coat Color

- Cattle: MC1R gene (red vs. black)
- Horses: Multiple loci (bay, chestnut, black, white patterns)

#### Polled/Horned Status

- Cattle: POLLED gene (polled is dominant)

#### Genetic Defects

- Swine: Halothane sensitivity (RYR1 gene)
- Cattle: Hypotrichosis (hairlessness)
- Cattle: Arthrogryposis, lethal recessives

#### Muscular Hypertrophy

- Cattle: Myostatin mutation (double muscling in Belgian Blue)
- Sheep: Myostatin mutation (Texel, Beltex)

### Breeding Approaches

[Content to be developed: Genotype and select/eliminate based on marker tests.]

## Quantitative Traits

[Content to be developed: Traits controlled by many genes, each with small effect.]

### Characteristics

[Content to be developed:]

- Polygenic inheritance (many loci)
- Continuous distribution
- Small effect per locus
- Strongly influenced by environment
- Require statistical methods for genetic evaluation

### Examples in Livestock

[Content to be developed:]

#### Growth and Feed Efficiency

- Average daily gain
- Feed conversion ratio
- Residual feed intake

#### Reproductive Performance

- Litter size in swine
- Fertility in cattle
- Age at puberty

#### Production Traits

- Milk yield and components
- Egg production
- Carcass composition

#### Health and Fitness

- Disease resistance
- Longevity
- Leg soundness

### Why Most Economic Traits are Quantitative

[Content to be developed: Complex traits like growth, reproduction, and production involve many physiological processes, each influenced by multiple genes.]

### Breeding Approaches

[Content to be developed: Estimate breeding values, apply selection based on EBVs, use genomic selection.]

## Threshold Traits

[Content to be developed: Traits that appear categorical but have an underlying continuous distribution.]

### Concept

[Content to be developed: Liability scale with threshold. When liability exceeds threshold, trait is expressed.]

### Examples

[Content to be developed:]

- Calving difficulty (normal vs. difficult)
- Teat number defects in swine (normal vs. inverted)
- Retained placenta in cattle

### Genetic Analysis

[Content to be developed: Can be analyzed on observed (0/1) scale or underlying liability scale.]

## Mixed Model Traits

[Content to be developed: Mostly quantitative with a major gene effect.]

### Example: Halothane in Swine

[Content to be developed: Growth and meat quality are quantitative, but RYR1 gene has major effect on stress susceptibility and meat quality.]

### Modern Genomics Approach

[Content to be developed: Genomic selection captures both polygenic and major QTL effects.]

## Comparing Simply Inherited and Quantitative Traits

[Content to be developed: Table comparing characteristics, examples, and breeding approaches.]

## R Demonstration: Simulating Different Trait Types

[Content to be developed:]

```{r}
#| echo: true
#| eval: false

# Simulate simply inherited trait (single locus, two alleles)
set.seed(42)
n <- 1000
allele1 <- sample(c("A", "a"), n, replace = TRUE, prob = c(0.7, 0.3))
allele2 <- sample(c("A", "a"), n, replace = TRUE, prob = c(0.7, 0.3))
genotype <- paste0(allele1, allele2)
phenotype_simple <- case_when(
  genotype %in% c("AA", "Aa", "aA") ~ "Dominant",
  genotype == "aa" ~ "Recessive"
)

# Simulate quantitative trait (100 loci, each with small effect)
n_loci <- 100
effects <- rnorm(n_loci, mean = 0, sd = 0.5)
genotype_matrix <- matrix(sample(0:2, n * n_loci, replace = TRUE), nrow = n, ncol = n_loci)
genetic_value <- genotype_matrix %*% effects
environmental_effect <- rnorm(n, mean = 0, sd = 2)
phenotype_quant <- genetic_value + environmental_effect

# Plot distributions
par(mfrow = c(1, 2))
barplot(table(phenotype_simple), main = "Simply Inherited Trait", ylab = "Frequency")
hist(phenotype_quant, main = "Quantitative Trait", xlab = "Phenotype", col = "lightblue")
```

## Summary

[Content to be developed.]

### Key Points

- Simply inherited traits: single gene, major effect, qualitative, can be fixed or eliminated
- Quantitative traits: many genes, small effects each, continuous distribution, gradual improvement
- Most economically important traits are quantitative
- Breeding strategies differ between trait types
- Modern genomics helps identify major genes within quantitative traits

## Practice Problems

[Problems to be developed]

1. Classify each trait as simply inherited or quantitative: (a) coat color in horses, (b) milk yield in dairy cattle, (c) polled status in cattle, (d) litter size in swine.

2. Explain why litter size in swine cannot be controlled by a single gene.

3. A breeder wants to eliminate a recessive lethal defect from a population. Describe a breeding strategy using molecular testing.

## Further Reading

[References to be added]

- Genetic defects in livestock
- QTL mapping studies
- Major genes affecting economic traits
