# Crossbreeding and Hybrid Vigor {#sec-crossbreeding}

<div class="learning-objectives">
### Learning Objectives {.unnumbered}

By the end of this chapter, you will be able to:

1. Define crossbreeding and hybrid vigor (heterosis)
2. Explain the genetic basis of heterosis
3. Describe which traits exhibit high vs. low heterosis
4. Compare different crossbreeding systems (terminal, rotational, composite)
5. Understand industry breeding structures by species

</div>

```{r}
#| echo: false
#| message: false
#| warning: false

# Load required packages
library(tidyverse)
library(knitr)

# Set default theme for plots
theme_set(theme_minimal(base_size = 12))
```

## Introduction

[Content to be developed: Crossbreeding is the mating of animals from different breeds or lines. It exploits hybrid vigor (heterosis) and breed complementarity to improve performance.]

## What is Crossbreeding?

[Content to be developed: Define crossbreeding.]

### Definition

[Content to be developed: Mating animals from different breeds, lines, or populations.]

### Why Crossbreed?

[Content to be developed:]

1. **Heterosis (hybrid vigor)**: Crossbreds outperform purebreds
2. **Breed complementarity**: Combine strengths of different breeds
3. **Faster genetic improvement**: In some systems, select within purebreds while producing crossbred commercial animals

### Species Where Crossbreeding is Common

[Content to be developed:]

- **Swine**: Universal (three-breed crosses)
- **Beef cattle**: Very common (crossbred cows, terminal sires)
- **Sheep**: Common (terminal sire systems)
- **Poultry**: Universal (proprietary crosses)
- **Dairy cattle**: Less common, mostly purebred Holsteins (some crossbreeding with Jersey, other breeds)
- **Horses**: Rare (breed registries require purebreds)

## Hybrid Vigor (Heterosis)

[Content to be developed: The performance advantage of crossbreds.]

### Definition

[Content to be developed:]

$$
\text{Heterosis} = \text{Performance}_{F1} - \text{Average performance of purebred parents}
$$

Often expressed as percentage:

$$
\text{Heterosis (\%)} = \frac{\text{Performance}_{F1} - \text{Avg purebred}}{\text{Avg purebred}} \times 100
$$

### Example

[Content to be developed:]

- Breed A litter size: 10.0 pigs
- Breed B litter size: 11.0 pigs
- F₁ cross (A × B) litter size: 11.5 pigs
- Midparent: (10.0 + 11.0) / 2 = 10.5
- Heterosis: 11.5 - 10.5 = 1.0 pig (9.5%)

## Genetic Basis of Heterosis

[Content to be developed: What causes hybrid vigor?]

### Dominance

[Content to be developed:]

- Crossbreds are more heterozygous (Aa) than purebreds (AA or aa)
- Dominant alleles mask recessive deleterious alleles
- Major contributor to heterosis

### Overdominance

[Content to be developed:]

- Heterozygote (Aa) is superior to both homozygotes (AA, aa)
- Rare, but may contribute to heterosis at some loci

### Epistasis

[Content to be developed:]

- Favorable interactions between alleles from different breeds
- Difficult to measure, but may contribute

## Traits with High vs. Low Heterosis

[Content to be developed: Heterosis varies by trait.]

### High Heterosis (Reproductive and Fitness Traits)

[Content to be developed:]

| Trait | Typical Heterosis (%) |
|-------|------------------------|
| Litter size (swine) | 5-15% |
| Piglet survival | 5-10% |
| Fertility (cattle) | 5-10% |
| Calf survival | 5-10% |
| Longevity | 10-20% |

### Moderate Heterosis (Growth and Efficiency)

[Content to be developed:]

| Trait | Typical Heterosis (%) |
|-------|------------------------|
| Average daily gain | 2-8% |
| Feed efficiency | 2-5% |
| Weaning weight | 3-8% |

### Low Heterosis (Carcass and Production Traits)

[Content to be developed:]

| Trait | Typical Heterosis (%) |
|-------|------------------------|
| Backfat thickness | 0-3% |
| Loin depth | 0-2% |
| Milk yield | 0-5% |
| Egg production | 0-3% |

### Why Fitness Traits Show High Heterosis

[Content to be developed: Fitness traits have been under natural selection for millennia. Deleterious recessive alleles persist at low frequency, and crossbreeding masks them.]

## Crossbreeding Systems

[Content to be developed: Different strategies for using crossbreeding.]

### Terminal Cross

[Content to be developed:]

- Purebred or F₁ females × terminal sire breed
- All offspring go to slaughter (none kept for breeding)
- Maximizes heterosis in market animals
- Common in beef (Angus, Charolais, Simmental sires on crossbred cows)

### Two-Breed Rotation

[Content to be developed:]

- Alternate breeds each generation
- Maintain ~67% heterosis after reaching equilibrium
- Example: Angus cows bred to Hereford bulls → F₁ females bred back to Angus bulls → repeat

### Three-Breed Rotation

[Content to be developed:]

- Rotate among three breeds
- Maintain ~86% heterosis
- Common in swine (e.g., Yorkshire, Landrace, Duroc)

### Composite Breeds

[Content to be developed:]

- Fixed breed composition (e.g., 50% Breed A, 50% Breed B)
- Maintain heterosis across loci that differ between breeds
- Example: Beefmaster (50% Brahman, 25% Hereford, 25% Shorthorn)

### Rotational-Terminal Cross

[Content to be developed:]

- Produce replacement females through rotation
- Breed market animals using terminal sire
- Combines heterosis in both females and market offspring

## Industry Structure by Species

[Content to be developed: How crossbreeding is implemented in commercial production.]

### Swine

[Content to be developed:]

- **F₁ females**: Large White × Landrace (maternal traits)
- **Terminal sire**: Duroc, Pietrain, or synthetic line (growth, carcass)
- **Three-way cross**: (LW × LR) F₁ female × Duroc sire → market pigs
- Nearly 100% of market pigs are three-way crosses

### Beef Cattle

[Content to be developed:]

- **Crossbred cows**: Various maternal breeds (Angus, Hereford, Red Angus)
- **Terminal sires**: Angus, Charolais, Simmental, Limousin
- Rotational or terminal systems common

### Poultry

[Content to be developed:]

- **Broilers**: Proprietary four-way crosses (great-grandparent → grandparent → parent → commercial)
- **Layers**: Proprietary crosses (White Leghorn lines or brown egg layer lines)
- Highly integrated, genetics controlled by 3-4 global companies

### Sheep

[Content to be developed:]

- **Crossbred ewes**: Various maternal breeds
- **Terminal sires**: Suffolk, Hampshire, Texel (meat production)

### Dairy Cattle

[Content to be developed:]

- Mostly purebred (Holstein, Jersey)
- Some crossbreeding: Holstein × Jersey, Holstein × Scandinavian Red
- Heterosis in fertility and health, but less in milk yield

## Genetic Evaluation in Crossbreeding Systems

[Content to be developed: How to predict crossbred performance.]

### Purebred vs. Crossbred Performance

[Content to be developed: Selection in purebreds must predict performance in crossbred commercial animals.]

### General Combining Ability (GCA)

[Content to be developed: Average performance of a breed in crosses.]

### Specific Combining Ability (SCA)

[Content to be developed: Performance of a specific breed combination (beyond GCA).]

### Modern Genomic Evaluations

[Content to be developed: Genomic selection can estimate crossbred performance from purebred and crossbred data.]

## R Demonstration: Calculating Heterosis

[Content to be developed:]

```{r}
#| echo: true
#| eval: false

# Calculate heterosis
breed_A <- 10.0   # Litter size
breed_B <- 11.0
F1_cross <- 11.5

midparent <- (breed_A + breed_B) / 2
heterosis_absolute <- F1_cross - midparent
heterosis_percent <- (heterosis_absolute / midparent) * 100

cat("Breed A:", breed_A, "\n")
cat("Breed B:", breed_B, "\n")
cat("F1 cross:", F1_cross, "\n")
cat("Midparent:", midparent, "\n")
cat("Heterosis (absolute):", round(heterosis_absolute, 2), "\n")
cat("Heterosis (%):", round(heterosis_percent, 2), "%\n")
```

## R Demonstration: Simulating Rotational Crossbreeding

[Content to be developed:]

```{r}
#| echo: true
#| eval: false

# Simulate breed composition in two-breed rotation
generations <- 0:10
breed_A_proportion <- numeric(length(generations))
breed_A_proportion[1] <- 1.0  # Start with purebred A

# Two-breed rotation: alternate between breeds
for (i in 2:length(generations)) {
  # Each generation, breed back to the other breed
  # Offspring get 50% from sire (the other breed) and 50% from dam
  if (i %% 2 == 0) {
    breed_A_proportion[i] <- 0.5 * breed_A_proportion[i-1]
  } else {
    breed_A_proportion[i] <- 0.5 + 0.5 * breed_A_proportion[i-1]
  }
}

# Plot
data.frame(Generation = generations, Breed_A = breed_A_proportion) %>%
  ggplot(aes(x = Generation, y = Breed_A)) +
  geom_line(color = "blue", size = 1.2) +
  geom_point(color = "blue", size = 2) +
  geom_hline(yintercept = 2/3, linetype = "dashed", color = "red") +
  labs(title = "Breed Composition in Two-Breed Rotation",
       x = "Generation",
       y = "Proportion of Breed A",
       caption = "Red dashed line: Equilibrium at 2/3") +
  theme_minimal()
```

## Summary

[Content to be developed.]

### Key Points

- Crossbreeding exploits heterosis (hybrid vigor) and breed complementarity
- Heterosis is highest for reproductive and fitness traits (5-20%)
- Terminal crosses maximize heterosis in market animals
- Rotational crosses maintain heterosis while producing replacement females
- Swine and poultry industries rely heavily on structured crossbreeding
- Genomic selection can predict crossbred performance from purebred data

## Practice Problems

[Problems to be developed]

1. Breed X has average daily gain of 0.80 kg/day. Breed Y has 0.85 kg/day. The F₁ cross has 0.86 kg/day. Calculate heterosis in absolute and percentage terms.

2. Explain why reproductive traits show much higher heterosis than carcass traits.

3. Compare two-breed rotation and three-breed rotation crossbreeding systems. What are the trade-offs?

4. Describe the three-way cross used in commercial swine production. Why is this system so widely used?

## Further Reading

[References to be added]

- Breed complementarity and heterosis estimates
- Papers on crossbreeding systems in swine, beef, poultry
